---
title: "MCP_Monarch"
format: html
editor: visual
---

## MCP Monarch

The Minimum Convex Polygon (MCP) is one of the simplest methods for estimating the home range or distributional range of a species. It works by drawing the smallest possible convex polygon that encloses a specified percentage of all observations. The 95% MCP represents the broad extent of the monarch's range, enclosing 95% of all recorded observations while excluding the most extreme outliers. The **50% MCP** represents the core use area, where monarchs are most frequently observed. In this analysis, the 95% MCP spans much of North America reflecting the monarch's well-known long-distance migration between the US/Canada and overwintering sites in Mexico, while the 50% MCP captures the core breeding region in the central and eastern United States.

```{r}
library(here) 
library(tidyverse) 
library(janitor)
library(adehabitatHR)
library(sp)
```

```{r}
# Load & join data 
monarch1 <- read.csv("~/Desktop/ESM211/data/Monarch1.csv")
monarch2 <- read.csv("~/Desktop/ESM211/data/Monarch2.csv")
monarch3 <- read.csv("~/Desktop/ESM211/data/observations.csv")
```

```{r}
monarch1 <- monarch1 %>% mutate(captive_cultivated = as.character(captive_cultivated),
                                 coordinates_obscured = as.character(coordinates_obscured))
monarch2 <- monarch2 %>% mutate(captive_cultivated = as.character(captive_cultivated),
                                 coordinates_obscured = as.character(coordinates_obscured))
monarch3 <- monarch3 %>% mutate(captive_cultivated = as.character(captive_cultivated),
                                 coordinates_obscured = as.character(coordinates_obscured))
```

```{r}
# Join data 
joined1and2 <- full_join(monarch1, monarch2)
joinedall <- full_join(joined1and2, monarch3)
```

```{r}
# Clean & filter 
joinedall <- clean_names(joinedall)

monarch_clean <- joinedall %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  filter(latitude >= -90 & latitude <= 90) %>%
  filter(longitude >= -180 & longitude <= 180)
```

```{r}
# Convert to spatial object 
coords <- monarch_clean %>%
  dplyr::select(longitude, latitude)

monarch_sp <- SpatialPointsDataFrame(
  coords = coords,
  data = monarch_clean,
  proj4string = CRS("+proj=longlat +datum=WGS84")
)
```

```{r}
# Add dummy ID column
monarch_clean$id <- as.factor("monarch")

# Convert to spatial object 
coords <- monarch_clean %>%
  dplyr::select(longitude, latitude)

monarch_sp <- SpatialPointsDataFrame(
  coords = coords,
  data = monarch_clean,
  proj4string = CRS("+proj=longlat +datum=WGS84")
)

# Run MCP 
monarch_mcp_95 <- mcp(monarch_sp[, "id"], percent = 95)
monarch_mcp_50 <- mcp(monarch_sp[, "id"], percent = 50)

print(monarch_mcp_95)
print(monarch_mcp_50)
```

```{r}
joinedall <- clean_names(joinedall)

monarch_clean <- joinedall %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  filter(latitude >= -90 & latitude <= 90) %>%
  filter(longitude >= -180 & longitude <= 180)

# Add dummy ID column
monarch_clean$id <- as.factor("monarch")

# Convert to spatial object 
coords <- monarch_clean %>%
  dplyr::select(longitude, latitude)

monarch_sp <- SpatialPointsDataFrame(
  coords = coords,
  data = monarch_clean,
  proj4string = CRS("+proj=longlat +datum=WGS84")
)

# Run MCP 
monarch_mcp_95 <- mcp(monarch_sp[, "id"], percent = 95)
monarch_mcp_50 <- mcp(monarch_sp[, "id"], percent = 50)

print(monarch_mcp_95)
print(monarch_mcp_50)

```

```{r}
# Plot
par(mar = c(4, 4, 3, 2))

plot(monarch_mcp_95, col = adjustcolor("mediumpurple1", alpha.f = 0.4), 
     border = "purple4", main = "Monarch Butterfly MCP Home Range",
     xlim = c(-170, -50), ylim = c(10, 75), asp = 1)
plot(monarch_sp, add = TRUE, pch = 1, cex = 0.3, col = "gray30")
plot(monarch_mcp_50, col = adjustcolor("darkorchid3", alpha.f = 0.5), 
     border = "purple4", add = TRUE)
legend("topright", legend = c("95% MCP", "50% MCP", "Observations"),
       fill = c(adjustcolor("mediumpurple1", alpha.f = 0.4), 
                adjustcolor("darkorchid3", alpha.f = 0.5), NA),
       pch = c(NA, NA, 1),
       col = c(NA, NA, "gray30"))
```

```{r}
# Subsample points for plotting
monarch_plot <- monarch_clean %>% slice_sample(n = 5000)

# Simple plot with fixed aspect ratio
par(mar = c(4, 4, 3, 2), bg = "aliceblue")

plot(monarch_mcp_95, col = adjustcolor("mediumpurple1", alpha.f = 0.4), 
     border = "purple4", lwd = 2,
     main = "Monarch Butterfly MCP Home Range",
     xlim = c(-130, -60), ylim = c(15, 60), asp = 1.3)

points(monarch_plot$longitude, monarch_plot$latitude, 
       pch = 16, cex = 0.3, col = adjustcolor("gray20", alpha.f = 0.4))

plot(monarch_mcp_50, col = adjustcolor("darkorchid3", alpha.f = 0.5), 
     border = "purple4", lwd = 2, add = TRUE)

legend("topright", legend = c("95% MCP", "50% MCP", "Observations"),
       fill = c(adjustcolor("mediumpurple1", alpha.f = 0.4), 
                adjustcolor("darkorchid3", alpha.f = 0.5), NA),
       pch = c(NA, NA, 16),
       col = c(NA, NA, "gray20"),
       bty = "n", cex = 0.9)
```

```{r}
# Subsample points for plotting
monarch_plot <- monarch_clean %>% slice_sample(n = 5000)

par(mar = c(4, 4, 3, 2), bg = "aliceblue")

plot(monarch_mcp_95, col = adjustcolor("mediumpurple1", alpha.f = 0.4), 
     border = "purple4", lwd = 2,
     main = "Monarch Butterfly MCP Home Range",
     xlim = c(-130, -60), ylim = c(15, 60), asp = 1.3)

points(monarch_plot$longitude, monarch_plot$latitude, 
       pch = 16, cex = 0.3, col = adjustcolor("gray20", alpha.f = 0.4))

plot(monarch_mcp_50, col = adjustcolor("darkorchid3", alpha.f = 0.5), 
     border = "purple4", lwd = 2, add = TRUE)

legend("topright", legend = c("95% MCP", "50% MCP", "Observations"),
       fill = c(adjustcolor("mediumpurple1", alpha.f = 0.4), 
                adjustcolor("darkorchid3", alpha.f = 0.5), NA),
       pch = c(NA, NA, 16),
       col = c(NA, NA, "gray20"),
       bty = "n", cex = 0.9)
```

```{r}
library(maps)

set.seed(42)
monarch_plot <- monarch_clean %>% slice_sample(n = 4000)

par(mar = c(4,4,3,1), bg = "white", xaxs = "i", yaxs = "i")

maps::map("world",
          regions = c("USA", "Canada", "Mexico"),
          fill = TRUE,
          col = "gray95",
          border = "gray70",
          xlim = c(-130, -60),
          ylim = c(15, 60))

plot(monarch_mcp_95,
     add = TRUE,
     col = adjustcolor("mediumpurple1", 0.18),
     border = adjustcolor("purple4", 0.5),
     lwd = 1)

plot(monarch_mcp_50,
     add = TRUE,
     col = adjustcolor("darkorchid3", 0.6),
     border = "purple4",
     lwd = 2.5)

points(monarch_plot$longitude,
       monarch_plot$latitude,
       pch = 16,
       cex = 0.2,
       col = adjustcolor("black", 0.25))

title(main = "Monarch Butterfly Distribution",
      cex.main = 1.5,
      font.main = 2)

mtext("Minimum Convex Polygon (95% = Broad Range, 50% = Core Use Area)",
      side = 3,
      line = -0.5,
      cex = 0.9)

legend("bottomleft",
       legend = c("95% MCP", "50% MCP", "Observations"),
       fill = c(adjustcolor("mediumpurple1", 0.18),
                adjustcolor("darkorchid3", 0.6),
                NA),
       pch = c(NA, NA, 16),
       col = c(NA, NA, adjustcolor("black", 0.5)),
       pt.cex = 0.8,
       bty = "n",
       cex = 0.9)
```

```{r}
library(maps)

set.seed(42)
monarch_plot <- monarch_clean %>% slice_sample(n = 4000)

# Calculate bounding box from 95% MCP
mcp_bbox <- bbox(monarch_mcp_95)  # returns xmin, xmax, ymin, ymax

par(mar = c(4,4,3,1), bg = "white", xaxs = "i", yaxs = "i")

# Draw base map
maps::map("world",
          regions = c("USA", "Canada", "Mexico"),
          fill = TRUE,
          col = "gray95",
          border = "gray70",
          xlim = c(mcp_bbox[1,1] - 5, mcp_bbox[1,2] + 5),
          ylim = c(mcp_bbox[2,1] - 5, mcp_bbox[2,2] + 5))

# Add MCP polygons with orange shades
plot(monarch_mcp_95,
     add = TRUE,
     col = adjustcolor("orange1", alpha.f = 0.25),  # lighter orange
     border = adjustcolor("darkorange2", 0.7),
     lwd = 1)

plot(monarch_mcp_50,
     add = TRUE,
     col = adjustcolor("darkorange", alpha.f = 0.5),  # darker orange
     border = "darkorange3",
     lwd = 2.5)

# Add observation points
points(monarch_plot$longitude,
       monarch_plot$latitude,
       pch = 16,
       cex = 0.2,
       col = adjustcolor("orangered", 0.3))  # subtle orange points

# Add title and subtitle
title(main = "Monarch Butterfly Distribution",
      cex.main = 1.5,
      font.main = 2)

mtext("Minimum Convex Polygon (95% = Broad Range, 50% = Core Use Area)",
      side = 3,
      line = -0.5,
      cex = 0.9)

# Add legend
legend("bottomleft",
       legend = c("95% MCP", "50% MCP", "Observations"),
       fill = c(adjustcolor("orange1", 0.25),
                adjustcolor("darkorange", 0.5),
                NA),
       pch = c(NA, NA, 16),
       col = c(NA, NA, adjustcolor("orangered", 0.3)),
       pt.cex = 0.8,
       bty = "n",
       cex = 0.9)
```

```{r}
library(ggplot2)

# Convert MCP to data frames
mcp_95_df <- fortify(monarch_mcp_95)
mcp_50_df <- fortify(monarch_mcp_50)

# Subsample points so it doesn't crash
monarch_plot <- monarch_clean %>% slice_sample(n = 10000)

# Get built-in world map from ggplot2
world <- map_data("world")

ggplot() +
  geom_polygon(data = world, aes(x = long, y = lat, group = group),
               fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_polygon(data = mcp_95_df, aes(x = long, y = lat, group = group),
               fill = adjustcolor("mediumpurple1", alpha.f = 0.3),
               color = "purple4", linewidth = 0.8) +
  geom_polygon(data = mcp_50_df, aes(x = long, y = lat, group = group),
               fill = adjustcolor("darkorchid3", alpha.f = 0.4),
               color = "purple4", linewidth = 0.8) +
  geom_point(data = monarch_plot, aes(x = longitude, y = latitude),
             size = 0.2, color = "gray20", alpha = 0.4) +
  coord_fixed(ratio = 1.3, xlim = c(-170, -50), ylim = c(10, 75)) +
  labs(title = "Monarch Butterfly MCP Home Range",
       subtitle = "2010-2023 iNaturalist Observations",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```
